---
title: "PcrPlsRegression"
author: "Chase Enzweiler"
date: "10/9/2017"
output: html_document
---

# PCR & PLS Regression

```{r}
# load hitters data
library("ISLR")
library("pls")

data_hitters <- Hitters

# principal component regression

pcr_fit <- pcr(Salary ~ ., data = data_hitters, scale = TRUE, validation = "none")
names(pcr_fit)


```

## 1.) Start with PCA

Remove observations from hitters that have missing values in salary

```{r}
# remove salary missing values

data_hitters <- data_hitters[complete.cases(data_hitters[c("Salary")]),]

```
\ 
use model.matrix() to create a design matrix based of the formula salary~.
\ 
```{r}
# create the design matrix

design_matrix <- model.matrix(Salary ~ ., data = data_hitters)

```
\ 
remove the intercept column from design matrix
\ 
```{r}
# remove the first column of ones

design_matrix <- design_matrix[,-1]

```
\ 
standardize the model matrix of responses to create X
\ 
```{r}
# standardize the design matrix
X <- scale(design_matrix)

```
\ 
Use svd() to get singular value decomposition of $X = UDV^T$
\ 
```{r}
# singular decomposition

decomp <- svd(X)

```
\ 
compute the principal components Z from the from the standardized X and the eigenvectors V
\ 
```{r}
# principal components

Z <- X %*% decomp$v

```
\ 
Confirm that you pc's match the pcr_fit$scores
\ 
```{r}
# check equivalence of both methods
round(pcr_fit$scores, digits = 4) == round(Z, digits = 4)

```
\ 
\ 

## 2. PC Regression on the first component
\ 
\ 

use the first component to compute regression y on $z_1$, so we obtain first pcr coefficient $b_1$ by $$b_1 = (z_1^T z_1)^{-1} z_1^T y$$ 
\ 
then compute predicted values y_hat = $b_1z_1$
\ 
compare your fitted values against the fitted values from pcr which is pcr_fit$fitted.values[, , 1] which is the fitted response using PC1 provided by pcr
\ 
\ 
```{r}
# response vector
Y <- data_hitters$Salary

# first pcr coefficient
b_1 <- solve(t(Z[,1]) %*% Z[,1]) %*% t(Z[,1]) %*% Y

# pcr fitted value for component 1

y_hat_comp1 <- b_1 * Z[,1]

# check with pcr fitted values 

round(y_hat_comp1 + mean(Y), digits = 4) == round(pcr_fit$fitted.values[, , 1], digits = 4)


```
\ 

\ 

## 3.) PC Regression on all PC's

\ 
\ 

Compute the vector of PCR Coefficients $b_{pcr}$ by regressing Y on all PC's Z $$b_{pcr} = (Z^T Z)^{-1} Z^T Y$$

```{r}
# vector of pcr coefficients

b_pcr <- solve(t(Z) %*% Z) %*% t(Z) %*% Y


```
\ 
compute the vector of predicted values y_hat using all PC's
$$y_{hat} = Zb_{pcr}$$
 \ 
```{r}
# vector of predicted values using all PC's

y_hat_allpc <- Z %*% b_pcr
 
```
\ 
compare computed fitted vaues with pcr, pcr_fit$fitted.values[, , 19]
\ 
```{r}
# check that they are equal

# changes with all PC's included, us all pc's up to 

round(y_hat_allpc + mean(Y), digits = 4) == round(pcr_fit$fitted.values[, , 19], digits = 4)

```

\ 
\ 

## PCR coefficients in terms of predictor variables
\ 
\ 
You can rewrite the regression of PC1 in terms of the response variables as: $ y_{hat} = X (b_1 v_1)$
\ 
where $v_1$ is the loading associated with the first PC aka the first column of V
\ 
also $b^{star}_1 = b_1 v_1$ is the vector of regression coefficients in terms of the predictors
\ 
\ 
In general, the PC regression coefficients can be expressed in terms of the predictors as: $$b^{star}_k = V_k D^{-1}_k U^{T}_k Y$$

Where the index k indicates matrices associated to the first k components
\ 
\ 
```{r}
# calculate b_1*

b_1star <- b_1 * decomp$v[,1]

# check that the vector matches coefficients from pcr

round(b_1star, digits = 4) == round(pcr_fit$coefficients[, , 1], digits = 4)

# now do the same for all possible sets of PC's and verify the coefficients against the output of pcr_fit$coefficients

# list will hold my coefficient vectors for each PC
star_coeff <- list()

# inverse of the diagonal singular values 
D_inverse <- solve(diag(decomp$d))

for (i in 1:19){
  
  star_coeff[[paste0("b_star", i)]] <- decomp$v[,1:i] %*% as.matrix(D_inverse[1:i,1:i]) %*% t(decomp$u)[1:i,] %*% Y
}

# now check that our generated coeficient vectors match the ones from pcr_fit$coefficients. Going to check the first and last as well as a few in between 

# coeficients up to 1st pc
round(star_coeff$b_star1, digits = 4) == round(pcr_fit$coefficients[, , 1], digits = 4)

# coefficient vector up to 5th pc
round(star_coeff$b_star5, digits = 4) == round(pcr_fit$coefficients[, , 5], digits = 4)

# coefficient vector up to 13th pc
round(star_coeff$b_star13, digits = 4) == round(pcr_fit$coefficients[, , 13], digits = 4)

# coefficient vector up to last pc
round(star_coeff$b_star19, digits = 4) == round(pcr_fit$coefficients[, , 19], digits = 4)




```
\ 
\ 

## Partial Least Squares Regression

\ 
\ 

```{r}
# run plsr 
# salary on our model matrix
# make sure to standardize Y
# because we are assuming X and Y are standardized


Y <- scale(Y)

plsr_fit <- plsr(Y ~ X)

```


Write the classic version of the PLSR algorithm

```{r}
library(Matrix)
# classic plsr algorithm

classic_plsr <- function(X, y){
  
  # performs partial least squares algorithm
  
  # [input] X standardized matrix of the predictors, 
  # Y standardized response variable
  
  # [Output] matrix of components, matrix of weights,
  # matrix of loadings, vector of coefficients, matrix of fitte values
  
  r <- rankMatrix(X)
  
  X_list <- list()
  
  y_list <- list()
  
  # loadings 
  p_list <- list()
  
  # normalized weights
  weights_list <- list()
  
  # coefficients 
  b_list <- list()
  
  # components
  Z_list <- list()
  
  # fitted values

  fitted_list <- list()    
 #=========================================================
  # this first part is as the first i in for loop because 
  # there is no indexing at zero
     
  w <- t(X) %*% y
    
  # normalize weight 
    
  weights_list[[1]] <- w / sqrt(sum(w^2))
  
  # compute Z 
  
  Z_list[[1]] <- (X %*% weights_list[[1]]) / as.numeric(t(weights_list[[1]]) %*% weights_list[[1]])
  
  # compute loadings
  
  p_list[[1]] <- (t(X) %*% Z_list[[1]]) / as.numeric(t(Z_list[[1]]) %*% Z_list[[1]])
  
  X_list[[1]] <- X - Z_list[[1]] %*% t(p_list[[1]])
  
  b_list[[1]] <- (t(y) %*% Z_list[[1]]) / as.numeric(t(Z_list[[1]]) %*% Z_list[[1]])
  
  y_list[[1]] <- y - as.numeric(b_list[[1]]) * Z_list[[1]]
  
  fitted_list[[1]] <- as.numeric(b_list[[1]]) * Z_list[[1]]
 
  
  #============================================= 
  # loop to calculate the rest for each PC
  for (i in 2:r){
    
    # calculate the weights/ direction
    # this is proportional to corr between X_j and y
    
    w <- t(X_list[[i - 1]]) %*% y_list[[i - 1]]
    
    weights_list[[i]] <- w / sqrt(sum(w^2))
    
    # calculate and store the component
    
    Z_list[[i]] <- (X_list[[i - 1]] %*% weights_list[[i]]) / as.numeric(t(weights_list[[i]]) %*% weights_list[[i]])
    
    # calculate the loadings and store
    
    p_list[[i]] <- (t(X_list[[i - 1]]) %*% Z_list[[i]]) / as.numeric(t(Z_list[[i]]) %*% Z_list[[i]])
    
    # recalculate and store X 
    
    X_list[[i]] <- X_list[[i - 1]] - (Z_list[[i]] %*% t(p_list[[i]]))
    
    # calculate coefficients
    
    b_list[[i]] <- (t(y_list[[i - 1]]) %*% Z_list[[i]]) / as.numeric(t(Z_list[[i]]) %*% Z_list[[i]])
    
    # recalculate response
    
    y_list[[i]] <- y_list[[i - 1]] - as.numeric(b_list[[i]]) * Z_list[[i]]
    
    # fitted values
    
    fitted_list[[i]] <- as.numeric(b_list[[i]]) * Z_list[[i]]
                                  
  }
  
  return(list(components = do.call(cbind, Z_list), weights = do.call(cbind, weights_list), loadings = do.call(cbind, p_list),
         coefficients = do.call(c, b_list), fitted = do.call(cbind, fitted_list)))
  }

```
\ 
\ 
Now check our first PLS component with the built in function plsr
\ 
\ 
```{r}
# my plsr algorithm

my_plsr <- classic_plsr(X, Y)

# check the z1 components scores

round(my_plsr$components[,1], digits = 4) == round(plsr_fit$scores[,1], digits = 4)

# check the weights

round(my_plsr$weights[,1], digits = 4) == round(plsr_fit$loading.weights[,1], digits = 4)

# check the loadings 

round(my_plsr$loadings[,1], digits = 4) == round(plsr_fit$loadings[,1], digits = 4)

# check the fitted values
round(my_plsr$fitted[,1], digits = 4) == round(plsr_fit$fitted.values[, ,1], digits = 4)



```




